---
title: "HW3_3"
author: "Ratkevich Ilya"
date: '21/05/24 '
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Let's tackle the given tasks using R code. We'll start by loading the data and necessary libraries, and then proceed to analyze the data according to the specified tasks.

### Loading Data and Libraries

```{r}
# Load necessary libraries
library(readxl)
library(ggplot2)
library(dplyr)
library(vegan)
library(psych)
library(ggpubr)

# Set a consistent theme for plots
theme_set(theme_bw())

# Load the dataset
lizard_data <- read_excel('Sleepy lizard.xlsx')

# Select relevant features and convert certain columns to factors
selected_data <- lizard_data %>%
  select(Treatment, Habitat, Connectivity, Tot_WBC, Het_ABS, Lym_ABS, `H:L Ratio`, Mon_ABS, OthG_ABS, LBSI) %>%
  mutate(
    Treatment = as.factor(Treatment),
    Habitat = as.factor(Habitat),
    Connectivity = as.factor(Connectivity)
  )
```

### Data Normalization

```{r}
# Subset data for blood composition
blood_composition <- selected_data %>%
  select(-Treatment, -Habitat, -Connectivity)

# Log normalization of blood composition data
normalized_blood_data <- scale(log(blood_composition + 1), scale = FALSE)
```

### Task 1a: Blood Composition in Unmodified vs. Heavily Modified Landscapes

```{r}
# PERMANOVA for Treatment
result_1a <- adonis2(normalized_blood_data ~ selected_data$Treatment, method = "euclidean")
print(result_1a)

# Boxplot for visual comparison
ggboxplot(selected_data, x = "Treatment", y = "Tot_WBC", color = "Treatment", add = "jitter") +
  labs(title = "Total WBC by Treatment", y = "Total WBC")
```

### Task 1b: Blood Composition in Different Habitats in Heavily Modified Landscapes

```{r}
# Subset data for Treatment 2
subset_treatment2 <- selected_data[selected_data$Treatment == 2, ]
subset_blood_treatment2 <- normalized_blood_data[selected_data$Treatment == 2, ]

# PERMANOVA for Habitat within Treatment 2
result_1b <- adonis2(subset_blood_treatment2 ~ subset_treatment2$Habitat, method = "euclidean")
print(result_1b)

# Boxplot for visual comparison
ggboxplot(subset_treatment2, x = "Habitat", y = "Tot_WBC", color = "Habitat", add = "jitter") +
  labs(title = "Total WBC by Habitat in Heavily Modified Landscape", y = "Total WBC")
```

### Task 1c: Blood Composition and Connectivity in Heavily Modified Landscapes

```{r}
# PERMANOVA for Connectivity within Treatment 2
result_1c <- adonis2(subset_blood_treatment2 ~ subset_treatment2$Connectivity, method = "euclidean")
print(result_1c)

# Boxplot for visual comparison
ggboxplot(subset_treatment2, x = "Connectivity", y = "Tot_WBC", color = "Connectivity", add = "jitter") +
  labs(title = "Total WBC by Connectivity in Heavily Modified Landscape", y = "Total WBC")
```

### Task 2: Combined Effects of Habitat Type and Connectivity

```{r}
# PERMANOVA for combined effects of Habitat and Connectivity within Treatment 2
result_2 <- adonis2(subset_blood_treatment2 ~ subset_treatment2$Habitat + subset_treatment2$Connectivity, method = "euclidean")
print(result_2)

# Interaction plot for visual comparison
ggboxplot(subset_treatment2, x = "Habitat", y = "Tot_WBC", color = "Connectivity", add = "jitter") +
  labs(title = "Total WBC by Habitat and Connectivity in Heavily Modified Landscape", y = "Total WBC") +
  facet_wrap(~ Connectivity)
```

### Explanation of Results

- **Task 1a**: The PERMANOVA analysis shows that the blood composition of lizards differs between unmodified and heavily modified landscapes.
- **Task 1b**: This analysis checks for differences in blood composition among lizards from different habitats within the heavily modified landscape.(not significant connection)
- **Task 1c**: This analysis tests the effect of the degree of connectivity on the blood composition of lizards within the heavily modified landscape.(not significant connection)
- **Task 2**: This analysis examines the combined effect of habitat type and degree of connectivity on the blood composition of lizards from the heavily modified landscape.(not significant connection)

### Visualization

```{r}
# Load additional libraries for enhanced visualization
library(ggforce)
library(ggrepel)

# Perform PCA
pca_result <- prcomp(normalized_blood_data)$x %>%
  as.data.frame() %>%
  select(PC1, PC2)

# Add treatment and habitat information to PCA result
pca_result$Treatment <- selected_data$Treatment
pca_result$Habitat <- selected_data$Habitat

# Plot PCA results with enhanced design
ggplot(pca_result, aes(x = PC1, y = PC2, shape = Treatment, color = Habitat)) +
  geom_point(size = 3, alpha = 0.8) +  # Increase point size and add transparency
  stat_ellipse(aes(fill = Treatment), geom = "polygon", alpha = 0.2, show.legend = FALSE) +  # Ellipses for group clustering
  geom_text_repel(aes(label = rownames(pca_result)), size = 3) +  # Add labels to points
  labs(
    title = "PCA of Blood Composition",
    subtitle = "Colored by Habitat and Shaped by Treatment",
    x = "Principal Component 1",
    y = "Principal Component 2"
  ) +
  scale_color_manual(values = c("blue", "green", "red", "orange")) +  # Custom colors for habitats
  scale_shape_manual(values = c(16, 17)) +  # Custom shapes for treatments
  theme_minimal(base_size = 15) +  # Minimal theme with larger base font size
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, face = "italic"),
    legend.position = "right",
    legend.title = element_blank(),
    legend.background = element_rect(fill = "white", color = "black")
  )

```


This helps to visualize how well the groups separate in the PCA space, providing further insights into the differences in blood composition among the different groups.

## Conclusion

The blood composition of lizards from unmodified and heavily modified landscapes differs
Blood composition of lizards in different habitats differs between lizards from the highly modified landscape doesn't differ much
The degree of connectivity doesn't affect the blood composition of lizards from the highly modified landscape 
The blood composition of lizards from a highly modified landscape isn't affected by habitat type and degree of connectivity 